---
title: "Titration of ERCC Spike"
author: "Jared Bard"
date: "25/06/30"
format:
  html:
    embed-resources: true
---

Sample is in vitro transcribed RNA from ERCC_013, prepared by Jeremy Pinheiro. Samples were made by making a dilution of RNA, starting from 1 ng/uL 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE) # options for how knitr displays html documents upon rendering

# first install packages if necessary
if (!requireNamespace("BiocManager")) install.packages("BiocManager")
if (!requireNamespace("tidyverse")) install.packages("tidyverse")
if (!requireNamespace("cowplot")) install.packages("cowplot")
if (!requireNamespace("ggrepel")) install.packages("ggrepel")
if (!requireNamespace("scales")) install.packages("scales")
if (!requireNamespace("ggokabeito")) install.packages("ggokabeito")
if (!requireNamespace("flextable")) install.packages("flextable")
if (!requireNamespace("readxl")) install.packages("readxl")
if (!requireNamespace("broom")) install.packages("broom")
if (!requireNamespace("devtools")) install.packages("devtools")
if (!requireNamespace("cat.extras")) devtools::install_github("jabard89/cat.extras") #convenient utility functions like prettier log scales

# then load packages
library(tidyverse)
library(cowplot)
library(cat.extras)
library(flextable)


conflicted::conflicts_prefer(dplyr::filter) # make sure that other packages don't take over "filter"
okabe_ito_colors <- palette.colors(n = NULL, palette = "Okabe-Ito", recycle = FALSE) #load okabe ito colorblind palette
`%!in%` = Negate(`%in%`)
theme_set(theme_half_open(font_size=10))
```

# reading in the data

```{r}
src.directory <- c(".")

samples_plate1 <- readxl::read_excel(file.path(src.directory,"250630_samples.xlsx"), sheet = "Plate1") %>%
  filter(!is.na(RNA_sample)) %>%
  mutate(Plate = "Plate1")
samples <- bind_rows(samples_plate1) # in case there are multiple plates of data

plate1 <- readxl::read_excel(file.path(src.directory,"250630-1.xls"), sheet = "Results", skip = 20) %>%
  mutate(Well = `Well Position`,
         Ct = as.numeric(CT)) %>%
  select(Well,Ct,Reporter) %>%
  right_join(samples_plate1 %>% filter(!is.na(Well)), by = "Well")

cycle_data <- readxl::read_excel(file.path(src.directory,"250630-1.xls"), sheet = "Multicomponent Data", skip = 20) %>%
  mutate(Well = `Well Position`) %>%
  select(Well,Cycle,ROX,FAM,VIC,TAMRA,CY5) %>%
  left_join(samples_plate1 %>% filter(!is.na(Well)), by = "Well") %>%
  pivot_longer(cols = c(ROX,FAM,VIC,TAMRA,CY5), names_to = "Channel", values_to = "Value") %>%
  # now go through and match probes to channels
  mutate(Probe = case_when(Channel == "FAM" ~ Probe_FAM,
                           Channel == "VIC" ~ Probe_SUN, #SUN and VIC are the same dye
                           Channel == "CY5" ~ Probe_Cy5))

raw_data <- bind_rows(plate1) %>% 
  replace_na(list(Skip=FALSE,RT=TRUE)) %>%
  # now go through and match probes to channels
  mutate(Probe = case_when(Reporter == "FAM" ~ Probe_FAM,
                           Reporter == "VIC" ~ Probe_SUN, #SUN and VIC are the same dye
                           Reporter == "CY5" ~ Probe_Cy5)) %>%
  filter(!is.na(Probe)) # remove any wells that don't have a probe

data_mean <- raw_data %>%
  filter(!Skip) %>%
  group_by(Plate,RNA_sample,Probe,RT) %>%
  summarise(Ct.SD = sd(Ct),
            Ct.SE = sd(Ct)/sqrt(n()),
            Ct.mean = mean(Ct)) %>%
  ungroup %>%
  left_join(samples %>% select(-c(Well,Techrep,RT,Skip)) %>% unique, by = c("Plate","RNA_sample"))
```
# have a peak at the raw raw data

```{r}
df <- cycle_data %>%
  separate(Well, into = c("Row", "Column"), sep = "(?<=\\D)(?=\\d)", remove = F) %>%
  mutate(Column = factor(as.numeric(Column))) %>%
  # filter(Row %in% LETTERS[which(LETTERS == "A"):which(LETTERS == "O")],
  #        Column %in% seq(2,17)) %>%
  filter(Row %in% c("B","C","E","F","H","I","K","L"),
         Column %in% seq(2,17)) %>%
  group_by(Plate,Well) %>%
  mutate(Value.norm = Value/Value[Channel == "ROX"]) %>%
  ungroup

ggplot(df %>% filter(Channel %in% c("FAM","CY5","VIC")),
       aes(x = Cycle, y= Value.norm, color = Channel)) +
  cowplot::theme_minimal_grid(font_size = 12) +
  geom_line() + 
  scale_y_log2nice() + 
  coord_cartesian(ylim = c(2^-3,2^4)) +
  facet_grid(Row ~ Column)

```
# first check the RT

```{r}
df <- raw_data

p_rt <-  ggplot(df,
               aes(x = RNA_sample, y = Ct, shape = Techrep)) + 
  cowplot::theme_minimal_grid(font_size = 12) +
  cowplot::panel_border() +
  geom_point(aes(color = RT)) +
  facet_grid(Probe ~ .) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p_rt
```

Generally technical replicates look good, though clearly some noise in hoNLuc_S1 (should be negative for all).

Now let's fit the data to a line of best fit and check the efficiency.

```{r}
if (!requireNamespace("ggpmisc")) install.packages("ggpmisc")
library(ggpmisc)
library(broom)   # For tidy() to easily extract model coefficients

# --- Step 1: Prepare and Analyze the Data ---

# Use the 'raw_data' tibble you have
# Filter for the specific probe and valid data points
ercc_data <- raw_data %>%
  filter(
    RT == TRUE,
    !is.na(Ct),
    Probe == "ERCC_013",
    RNA_conc_ng_per_uL > 0 # log(0) is undefined, so ensure concentration is positive
  )

# Fit the linear model outside of ggplot to get the slope
# We model Ct against the log10 of the concentration
model <- lm(Ct ~ log10(RNA_conc_ng_per_uL), data = ercc_data)

# Extract the slope using broom::tidy()
# The slope is the 'estimate' for the term 'log10(RNA_conc_ng_per_uL)'
slope <- tidy(model) %>%
  filter(term == "log10(RNA_conc_ng_per_uL)") %>%
  pull(estimate)

# Calculate the efficiency using the formula
efficiency <- (10^(-1/slope) - 1) * 100

# Create a text label for the plot annotation
# The `\n` creates a new line to separate the text from the equation
efficiency_label <- paste("\nEfficiency:", round(efficiency, 1), "%","\nIdeal is 90%-110%")


# --- Step 2: Create the ggplot ---

# Note: The function `scale_x_log10nice()` is from the 'ggprism' package.
# I am using the standard `scale_x_log10()` from ggplot2 for broader compatibility.
# If you have ggprism installed, you can use your original function.
qpcr_plot <- ggplot(
  data = ercc_data,
  aes(x = RNA_conc_ng_per_uL, y = Ct) # Color is not needed if only one probe is plotted
) +

  # Add points for the raw data
  geom_point(alpha = 0.8, size = 3, color = "firebrick") +

  # Use a logarithmic scale for the x-axis
  scale_x_log10(
    breaks = scales::trans_breaks("log10", function(x) 10^x),
    labels = scales::trans_format("log10", scales::math_format(10^.x)),
    name = "RNA Concentration (ng/µL)"
  ) +
  
  # Add the line of best fit (linear model 'lm')
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x, color = "black") +

  # Use stat_poly_eq to add the regression equation and R-squared value
  stat_poly_eq(
    formula = y ~ x,
    aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~")),
    parse = TRUE,
    label.x = "middle", # Position horizontally
    label.y = "top" # Position vertically
  ) +
  
  # --- Add the Efficiency Annotation ---
  # Use annotate() to add the calculated efficiency text to a specific spot
  ggplot2::annotate(
    "text",
    x = max(ercc_data$RNA_conc_ng_per_uL), # Position at the far right
    y = max(ercc_data$Ct) - 1,            # Position just below the regression equation
    label = efficiency_label,
    hjust = 1, # Right-align the text
    vjust = 1,
    size = 4,
    fontface = "bold"
  ) +
  
  # Improve labels and theme
  labs(
    title = "qPCR Standard Curve: ERCC_013",
    subtitle = "Ct Value vs. RNA Concentration",
    y = "Ct Value"
  ) +
  theme_bw() +
  theme(legend.position = "none") # Hide legend as we only have one series


# --- Step 3: Display the Plot ---
print(qpcr_plot)

# You can also print the calculated values to the console
cat("Slope:", slope, "\n")
cat("Efficiency:", efficiency, "%\n")
```

