---
title: "qPCR Viral Infection"
author: "Thu Nguyen"
date: "11/08/2024"
format:
  html:
    embed-resources: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE) # options for how knitr displays html documents upon rendering

# first install packages if necessary
if (!requireNamespace("BiocManager")) install.packages("BiocManager")
if (!requireNamespace("tidyverse")) install.packages("tidyverse")
if (!requireNamespace("cowplot")) install.packages("cowplot")
if (!requireNamespace("ggrepel")) install.packages("ggrepel")
if (!requireNamespace("scales")) install.packages("scales")
if (!requireNamespace("ggokabeito")) install.packages("ggokabeito")
if (!requireNamespace("flextable")) install.packages("flextable")
if (!requireNamespace("readxl")) install.packages("readxl")
if (!requireNamespace("devtools")) install.packages("devtools")
if (!requireNamespace("cat.extras")) devtools::install_github("jabard89/cat.extras") #convenient utility functions like prettier log scales

# then load packages
library(tidyverse)
library(cowplot)
library(cat.extras)
library(flextable)


conflicted::conflicts_prefer(dplyr::filter) # make sure that other packages don't take over "filter"
okabe_ito_colors <- list(
  black = "#000000",
  orange = "#E69F00", 
  sky_blue = "#56B4E9",
  bluish_green = "#009E73",
  yellow = "#F0E442",
  blue = "#0072B2",
  vermillion = "#D55E00",
  reddish_purple = "#CC79A7"
)
`%!in%` = Negate(`%in%`)
theme_set(theme_half_open(font_size=10))
```

# reading in the data

```{r}
src.directory <- c(".")

samples_plate1 <- readxl::read_excel(file.path(src.directory,"qPCR_241108_samples.xlsx"), sheet = "Plate1") %>%
  filter(!is.na(RNA_sample)) %>%
  mutate(Plate = "Plate1")
samples <- bind_rows(samples_plate1) # in case there are multiple plates of data

# this is the real data
plate1 <- readxl::read_excel(file.path(src.directory,"2024-11-08_ThuN.xls"), sheet = "Results", skip = 20) %>%
  mutate(Well = `Well Position`,
         Ct = as.numeric(CT)) %>%
  select(Well,Ct,Reporter) %>%
  right_join(samples_plate1 %>% filter(!is.na(Well)), by = "Well")

# this is the raw data for QC purposes
cycle_data <- readxl::read_excel(file.path(src.directory,"2024-11-08_ThuN.xls"), sheet = "Multicomponent Data", skip = 20) %>%
  mutate(Well = `Well Position`) %>%
  select(Well,Cycle,ROX,FAM,VIC,TAMRA,CY5) %>%
  left_join(samples_plate1 %>% filter(!is.na(Well)), by = "Well") %>%
  pivot_longer(cols = c(ROX,FAM,VIC,TAMRA,CY5), names_to = "Channel", values_to = "Value") %>%
  # now go through and match probes to channels
  mutate(Probe = case_when(Channel == "FAM" ~ Probe_FAM,
                           Channel == "TAMRA" ~ Probe_TAMRA,
                           Channel == "VIC" ~ Probe_SUN, #SUN and VIC are the same dye
                           Channel == "CY5" ~ Probe_Cy5))

raw_data <- bind_rows(plate1) %>% 
  replace_na(list(Skip=FALSE,RT=TRUE)) %>%
  # now go through and match probes to channels
  mutate(Probe = case_when(Reporter == "FAM" ~ Probe_FAM,
                           Reporter == "TAMRA" ~ Probe_TAMRA,
                           Reporter == "VIC" ~ Probe_SUN, #SUN and VIC are the same dye
                           Reporter == "CY5" ~ Probe_Cy5)) %>%
  filter(!is.na(Probe)) # remove any wells that don't have a probe

data_mean <- raw_data %>%
  filter(!Skip) %>%
  group_by(Plate,RNA_sample,Probe,RT) %>%
  summarise(Ct.SD = sd(Ct),
            Ct.SE = sd(Ct)/sqrt(n()),
            Ct.mean = mean(Ct)) %>%
  ungroup %>%
  left_join(samples %>% select(-c(Well,Techrep,RT,Skip), -starts_with("Probe_")) %>% unique, by = c("Plate","RNA_sample"))
```
# have a peak at the raw raw data

```{r}
df <- cycle_data %>%
  separate(Well, into = c("Row", "Column"), sep = "(?<=\\D)(?=\\d)", remove = F) %>%
  mutate(Column = factor(as.numeric(Column))) %>%
  filter(Row %in% LETTERS[which(LETTERS == "A"):which(LETTERS == "O")],
         Column %in% seq(1,24)) %>%
  group_by(Plate,Well) %>%
  mutate(Value.norm = Value/Value[Channel == "ROX"]) %>%
  ungroup

ggplot(df %>% filter(Channel %in% c("FAM","TAMRA","CY5","VIC")),
       aes(x = Cycle, y= Value.norm, color = Channel)) +
  cowplot::theme_minimal_grid(font_size = 12) +
  geom_line() + 
  scale_y_log2nice() + 
  coord_cartesian(ylim = c(2^-3,2^4)) +
  facet_grid(Row ~ Column)

```
# first check the RT

```{r}
df <- raw_data

p_rt <-  ggplot(df,
               aes(x = RNA_sample, y = Ct, shape = Techrep)) + 
  cowplot::theme_minimal_grid(font_size = 12) +
  cowplot::panel_border() +
  geom_point(aes(color = RT)) +
  facet_grid(Probe ~ .) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p_rt
```

Generally technical replicates look super tight, but C2_4-IFIT1-T1 looks off. We'll skip that well (I marked it as Skip on the sample sheet).

```{r}

df %>% filter(RNA_sample == "C2_4", RT == TRUE) %>% arrange(Probe,Techrep) %>%
  select(Well,RNA_sample,Probe,Techrep,Ct) %>% flextable %>% autofit
```

We are also missing a bunch of IFIT1 datapoints, probably because the signal was too low.

```{r}
df <- cycle_data %>%
  separate(Well, into = c("Row", "Column"), sep = "(?<=\\D)(?=\\d)", remove = F) %>%
  mutate(Column = factor(as.numeric(Column))) %>%
  filter(Row %in% LETTERS[which(LETTERS == "B"):which(LETTERS == "I")],
         Column %in% seq(2,17)) %>%
  group_by(Plate,Well) %>%
  mutate(Value.norm = Value/Value[Channel == "ROX"]) %>%
  ungroup %>%
  filter(RT == TRUE, Probe == "IFIT1_S1")

ggplot(df,
       aes(x = Cycle, y= Value.norm, color = FAZ3532_50uM)) +
  cowplot::theme_minimal_grid(font_size = 12) +
  ggokabeito::scale_color_okabe_ito() +
  geom_line(size = 1, aes(group = interaction(RNA_sample,Techrep))) + 
  scale_y_log2nice() + 
  coord_cartesian(ylim = c(2^-4,2^0)) +
  facet_grid(Virus_MOI ~ Time_hr)
```

# calculating relative abundances

Use the ∆∆Ct method to calculate the relative abundance of transcripts across conditions, normalized to a reference gene (GAPDH here):
https://www.nature.com/articles/nprot.2008.73.

Data is normalized to Time = 4hr, Virus_MOI = 1, FAZ3532_50uM = FALSE.

```{r}

# ----------------------------
# 0) Parameters
# ----------------------------
ref_probe      <- "GAPDH_S1"
ref_time       <- 4
ref_treatment  <- FALSE
ref_moi        <- 1

# ----------------------------
# 1) Keep only RT(+) rows and essential columns
# ----------------------------
df_rt <- data_mean %>%
  filter(RT == TRUE) %>%
  select(Bio_Sample, RNA_sample, Probe, Time_hr, FAZ3532_50uM, Virus_MOI, Ct.mean, Ct.SE)

# ----------------------------
# 2) Reference CONDITION values per Probe
#    (Time = ref_time, FAZ=ref_treatment, MOI=ref_moi)
# Computes mean of all biological samples at that condition
# ----------------------------
ref_cond_by_probe <- df_rt %>%
  filter(Time_hr == ref_time,
         FAZ3532_50uM == ref_treatment,
         Virus_MOI == ref_moi) %>%
  group_by(Probe) %>%
  summarise(
    Ct.refCond    = mean(Ct.mean, na.rm = TRUE),
    Ct.refCond.se = sqrt(sum((Ct.SE %>% replace_na(0))^2)) / sqrt(n()),
    .groups = "drop"
  )

# ----------------------------
# 3) Reference PROBE’s value under the reference condition
#    (a single-row tibble holding constants for all rows)
# ----------------------------
ref_probe_constants <- ref_cond_by_probe %>%
  filter(Probe == ref_probe) %>%
  transmute(
    Ct.refCond.refProbe    = Ct.refCond,
    Ct.refCond.refProbe.se = Ct.refCond.se
  )

# If the reference probe is missing, you may want to stop with a clear error:
# stopifnot(nrow(ref_probe_constants) == 1)

# ----------------------------
# 4) Reference PROBE per RNA sample
#    (housekeeper measured in the same sample)
# ----------------------------
ref_probe_by_sample <- df_rt %>%
  filter(Probe == ref_probe) %>%
  select(RNA_sample,
         Ct.refProbe    = Ct.mean,
         Ct.refProbe.se = Ct.SE) %>%
  distinct()

# ----------------------------
# 5) Build the analysis table by merging everything
# ----------------------------
df_merged <- df_rt %>%
  # add reference-condition stats for the matching probe
  left_join(ref_cond_by_probe, by = "Probe") %>%
  # add reference-probe value for the matching biological sample
  left_join(ref_probe_by_sample, by = "RNA_sample") %>%
  # cross-join the single-row constants so they're available to every row
  cross_join(ref_probe_constants)

# ----------------------------
# 6) Final calculations (ΔΔCt, Rel. abundance, and SE propagation)
# ----------------------------
df_out <- df_merged %>%
  mutate(
    # ΔCt for each row vs. the ref probe measured in the same sample
    delta_ct = Ct.mean - Ct.refProbe,

    # ΔCt for the reference condition (probe vs. ref probe under ref condition)
    delta_ct_refCond = Ct.refCond - Ct.refCond.refProbe,

    # ΔΔCt and relative abundance
    delta_delta_ct = delta_ct - delta_ct_refCond,
    Rel_abund      = 2^(-delta_delta_ct),

    # SE propagation
    delta_ct_se = sqrt((Ct.SE %>% replace_na(0))^2 +
                       (Ct.refProbe.se %>% replace_na(0))^2),

    delta_ct_refCond_se = sqrt((Ct.refCond.se %>% replace_na(0))^2 +
                               (Ct.refCond.refProbe.se %>% replace_na(0))^2),

    delta_delta_ct_se = sqrt(delta_ct_se^2 + delta_ct_refCond_se^2),
    Rel_abund_se      = Rel_abund * log(2) * delta_delta_ct_se
  )

# df_out is your final tibble with everything computed.

ggplot(df_out %>% filter(Probe == "CHIKV_E1"),
       aes(x = factor(Time_hr), y = Rel_abund, color = FAZ3532_50uM)) +
  geom_point() +
  stat_summary(fun = mean, geom = "crossbar", width = 0.5) +
  ggokabeito::scale_color_okabe_ito() +
  scale_y_log10nice() +
  facet_grid(Probe ~ Virus_MOI, labeller = labeller(Virus_MOI = function(x) paste("MOI", x))) +
  labs(x = "Time (hours)")
```

